---
format: 
  revealjs:
    slide-number: true
    # code-link: true
    highlight-style: a11y
    chalkboard: true
    theme: 
      - ../meds-slides-styles.scss
editor_options: 
  chunk_output_type: console
---

## {#title-slide data-menu-title="Title Slide" background="#053660"} 

[EDS 240: Lecture 3.2]{.custom-title}

[*Visualizing Distributions*]{.custom-subtitle}

<hr class="hr-teal">

[Week 3 | January 22^nd^, 2024]{.custom-subtitle3}

<!-- --- -->

<!-- ## {#distributions data-menu-title="Visualizing distributions" background="#047C90"}  -->

<!-- <div class="page-center vertical-center"> -->
<!-- <p class="custom-subtitle bottombr">Visualizing Distributions</p> -->
<!-- <p class="caption-text">boxplots, violin plots, histograms, density charts, ridgeline plots</p> -->
<!-- </div> -->

---

## {#viz-dist data-menu-title="Visualizing distributions"} 

[Visualizing distributions]{.slide-title}

<hr>

histograms
density plots  
box plots  
violin plots  
ridgeline plots

---

## {#hist-overview data-menu-title="Histogram overview"} 

[Histograms]{.slide-title}

<hr>

**What are they?** 

- Histograms are used to represent the distribution of a **numeric** variable(s), which is **cut into several bins**. The number of observations per bin is represented by the height of the bar.

**Need:**

- a numeric variable(s) with lots of values
- meaningful differences between values

**Important considerations:**

- bin width
- number of groups that you're comparing

---

## {#density-overview data-menu-title="Density plot overview"} 

[Density plots]{.slide-title}

<hr>

**What are they?**

- A smoothed version of a histogram. Density plots are representations of the distribution of a **numeric** variable(s), which uses a **[kernel density estimate](https://chemicalstatistician.wordpress.com/2013/06/09/exploratory-data-analysis-kernel-density-estimation-in-r-on-ozone-pollution-data-in-new-york-and-ozonopolis/)** to show the probability density function of the variable.

**Need:**

- a numeric variable(s) with lots of values

**Important considerations:**

- can be misleading with small data sets
- binsize(?)
- number of groups that you're comparing

---



---

## {#dist-numeric data-menu-title="Distribution of numeric vars"} 

[Distribution of a numeric variable(s)]{.slide-title2}

<hr>

::: panel-tabset

## Histogram

Histograms are used to represent the distribution of a **numeric** variable(s), which is **cut into several bins**. The number of observations per bin is represented by the height of the bar.

```{r}
#| eval: true
#| echo: true
#| out-width: "50%"
library(lterdatasampler)
library(tidyverse)

ggplot(knz_bison, aes(x = animal_weight, fill = animal_sex)) +
  geom_histogram(position = "identity", alpha = 0.7)
```

## Density plot

Density plots are representations of the distribution of a **numeric** variable(s), which uses a **[kernel density estimate](https://chemicalstatistician.wordpress.com/2013/06/09/exploratory-data-analysis-kernel-density-estimation-in-r-on-ozone-pollution-data-in-new-york-and-ozonopolis/)** to show the probability density function of the variable.

```{r}
#| eval: true
#| echo: true
#| out-width: "50%"
library(lterdatasampler)
library(tidyverse)

ggplot(knz_bison, aes(x = animal_weight, fill = animal_sex)) +
  geom_density(alpha = 0.7)
```

:::

---

## {#dist-num-considerations data-menu-title="Considerations (hist & dens)"} 

[Considerations]{.slide-title2}

<hr>

::: panel-tabset

## Bin width

**Try several bin widths, and make sure all bins are the same size.** By default, `geom_histogram()` uses 30 `bins`. Alternatively, adjust `binwidth`.

:::: columns

::: {.column width="50%"}
[` binwidth = 50`:]{.body-text-s}
```{r}
#| eval: true
#| echo: true
#| out-width: "90%"
library(lterdatasampler)
library(tidyverse)

ggplot(knz_bison, aes(x = animal_weight, fill = animal_sex)) +
  geom_histogram(position = "identity", alpha = 0.7, binwidth = 50)
```
:::

::: {.column width="50%"}
[`binwidth = 500`:]{.body-text-s}
```{r}
#| eval: true
#| echo: true
#| out-width: "90%"
library(lterdatasampler)
library(tidyverse)

ggplot(knz_bison, aes(x = animal_weight, fill = animal_sex)) +
  geom_histogram(position = "identity", alpha = 0.7, binwidth = 500)
```
:::

::::

## Group number

**Comparing more than ~3 groups in the same histogram leads to cluttered / hard to read plots.** Instead, try a violin/box/ridgeline plot or faceting:

:::: columns

::: {.column width="50%"}
[Too many groups...]{.body-text-s}
```{r}
#| eval: true
#| echo: true
#| out-width: "90%"
library(lterdatasampler)
library(tidyverse)

ggplot(pie_crab, aes(x = size, fill = site)) +
  geom_histogram(position = "identity", alpha = 0.7)
```
:::

::: {.column width="50%"}
[Also beware of unnecessary colors!]{.body-text-s}
```{r}
#| eval: true
#| echo: true
#| out-width: "90%"
library(lterdatasampler)
library(tidyverse)

ggplot(pie_crab, aes(x = size)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  facet_wrap(~site)
```
:::

::::

:::

---

<!-- STACKED DENSITY CHARTS -->
<!-- MIRRORED DENSITY / HISTOGRAM -->

<!-- ## {#variants-num-alt data-menu-title="Variants (hist & dens)"}  -->

<!-- [Variants]{.slide-title3} -->

<!-- <hr> -->

<!-- ::: panel-tabset -->

<!-- ## Mirrorred -->

<!-- ```{r} -->
<!-- library(lterdatasampler) -->
<!-- library(tidyverse) -->

<!-- # ggplot(knz_bison, aes(x = animal_weight, fill = animal_sex)) + -->
<!-- #   geom_histogram(position = "identity", alpha = 0.7, binwidth = 500) -->

<!-- female_bison <- knz_bison |> filter(animal_sex == "F") -->
<!-- male_bison <- knz_bison |> filter(animal_sex == "M") -->

<!-- # bison_wrangled <- knz_bison |>  -->
<!-- #   mutate(animal_weight = ifelse(test = grepl("F", animal_sex) == TRUE,  -->
<!-- #                                 yes = animal_weight,  -->
<!-- #                                 no = animal_weight * -1)) -->

<!-- ggplot(aes(x = x)) +  -->
<!--   geom_histogram(male_bison, aes(x = animal_weight, y = ..density.., fill = "#69b3a2")) -->

<!-- ggplot(data, aes(x=x) ) + -->
<!--   geom_histogram( aes(x = var1, y = ..density..), fill="#69b3a2" ) + -->
<!--   geom_label( aes(x=4.5, y=0.25, label="variable1"), color="#69b3a2") + -->
<!--   geom_histogram( aes(x = var2, y = -..density..), fill= "#404080") + -->
<!--   geom_label( aes(x=4.5, y=-0.25, label="variable2"), color="#404080") + -->
<!--   theme_ipsum() + -->
<!--   xlab("value of x") -->
<!-- # ggplot() + -->
<!-- #   geom_histogram(male_bison, mapping = aes(x = animal_weight)) +  -->
<!-- #   geom_histogram(female_bison, mapping = aes(x = -(animal_weight), fill = "blue")) -->
<!-- ``` -->


<!-- ## Marginal plots -->

<!-- ::: -->

<!-- --- -->

