---
title: "Lecture 1.3 KEY"
subtitle: "`{ggplot2}` review"
author: "Your Name"
date: 2025-01-06
toc: true
---

::: {.callout-note}
This template follows [lecture 1.3 slides](https://samanthacsik.github.io/EDS-240-data-viz/course-materials/lecture-slides/week1.3-ggplot-review-slides.html#/title-slide){target="_blank}.
:::

## Setup

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
# load packages ----
library(tidyverse)
library(palmerpenguins)
```

## Tidy Data Review

Example untidy / wide data: 

```{r}
# create some untidy temperature data ----
temp_data_wide <- tribble(
  ~date, ~station1, ~station2,  ~station3,
  "2023-10-01", 30.1, 29.8,  31.2,
  "2023-11-01", 28.6, 29.1,  33.4,
  "2023-12-01", 29.9, 28.5,  32.3
)

# print it out ----
print(temp_data_wide)
```

Using `pivot_longer()` to "lengthen" / tidy our data:

```{r}
# convert data from wide > long ----
temp_data_long <- temp_data_wide |> 
  pivot_longer(cols = starts_with("station"),
               names_to = "station_id",
               values_to = "temp_c")

# print it out ----
print(temp_data_long)
```

## Plot #1

Explore the relationship between penguin **bill length** and **bill depth**. Our goals are to review:

1. initializing a plot object and adding a geometry layer to represent our data
2. when to define data & map variables globally (e.g. within `ggplot()`) vs. locally (e.g. within a `geom_*()`)
3. updating how aesthetic mappings manifest visually aka scaling
4. piping into a ggplot

### Example 1 (the basics)

- initialize a plot object
- map aesthetics
- show that you can omit argument names
- add a geometry layer
- update how aesthetic mappings manifest visually (i.e. scaling)

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 2 (mapping custom colors)

- map custom colors rather than defaults

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 3 (mapping color to continuous variable)

- map color to a continuous variable (rather than categorical) 

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 4 (updating color for all points)

- color all points the same (i.e. not by a variable, like `species` or `body_mass_g`)

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 5 (local vs. global data & variable mappings) 

- define data and mappings within `geom_*()` (locally) rather than in `ggplot()` (globally) -- helpful if you plan to include multiple geoms with different mappings (e.g. you're plotting data from multiple data frames):

```{r}
#| warning: false
#| message: false
# create a separate penguins_summary df ----
penguins_summary <- penguins |>
  drop_na() |> 
  group_by(species) |> 
  summarize(
    mean_bill_length_mm = mean(bill_length_mm),
    mean_bill_depth_mm = mean(bill_depth_mm)
  )

# create ggplot with layers from different dfs ----
# ~ CODE ALONG HERE ~
```

### Example 6 & 7 (mapping variables globally vs. locally)

- global mappings are passed down to all subsequent layers:

```{r}
#| warning: false
#| message: false
# scatterplot with lms fitted to species ---- 
# ~ CODE ALONG HERE ~
```

- local mappings only apply to that particular layer:

```{r}
#| warning: false
#| message: false
# scatterplot with lm fitted to entire data set ---- 
# ~ CODE ALONG HERE ~
```

### Example 8 (piping data into a ggplot)

- pipe data into `ggplot()` (rather than specifying in the `data` argument) -- useful if you need to do a small bit of wrangling first, but don't want to create a whole new df

```{r}
#| warning: false
# filter for Gentoo penguins, then pipe into ggplot ----
# ~ CODE ALONG HERE ~
```

## Plot #2

Explore **penguin species counts**. Our goals are to review:

1. statistical transformations -- what are they, how to identify the default, and how to update them
2. position adjustments -- what are they, how to identify the default, and how to update them
3. coordinate systems -- what are they, how to identify the default, and how to update them    
4. updating non-data elements using pre-built themes and the `theme()` function

### Example 1 (the basics + default stat)

- initialize a plot object
- map aesthetics
- add a geometry layer
- explore default statistical transformations

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

- all geoms have a default statistical transformation
  - some geoms, like scatterplots, plot the raw values of your data set (e.g. `stat = "identity"`) 
  - other geoms, like bar chars, histograms, boxplots, etc. calculate *new* values to plot (e.g. `stat = count`)
  - reading documentation is important here!

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 2 (override default stat)

- let's say we have a df that already contains calculated species count values, which we want to plot:

```{r}
#| warning: false

# new summarized df ----
# ~ CODE ALONG HERE ~

# generate bar heights based on "identity" of values in `n` column ----
# ~ CODE ALONG HERE ~
```

### Example 3 (override default stat *mapping*)

- e.g. we can display the same data but with y-axis values as *proportions* rather than counts

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 4 (default position adjustment)

- position adjustments tweak position of elements to resolve overlapping geoms
- all geoms have a default position (e.g. barplots have `position = "stack"`) -- see documentation
- let's say we now want to visualize penguin counts by species (bar height) *and* by island (color):

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 5 & 6 (override default position adjustments)

- `position = "fill"` creates stacked bars of the same height (easier to compare proportions):

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

- `position = "dodge"` places overlapping bars directly beside one another (easier to compare individual values):


```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 7 (alternatively, use `position_*()`)

- instead of `position = "X"`, you can use a `position_*()` function to fine-tune position adjustments -- e.g. `position_dodge2()` to create equal width bars:

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 8 (default coordinate system)

- by default, ggplots are constructed in a Cartesian coordinate system

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 9 & 10 (alternative coordinate systems)

- flip x & y axes:

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

- polar coordinates: 

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 11 (pre-made themes)

- there are a number of built-in themes + extension packages

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

### Example 12 (customize further with `theme()`)

- modify any non-data element of a plot using `theme()`

```{r}
#| warning: false
# ~ CODE ALONG HERE ~
```

## Plot #3 

Explore **penguin flipper lengths**. Our goals are to review:

1. more position adjustments and scales
2. updating plot labels
3. faceting

### Example 1 (all in one go!)

- initialize a plot object
- map aesthetics
- add a geometry layer
- map color to species
- update position adjustment
- update labels
- facet by species

```{r}
#| warning: false
#| message: false
# ~ CODE ALONG HERE ~
```
