---
title: "Lecture 2.2 TEMPLATE"
subtitle: "Visualizing distributions"
author: "Your Name"
date: 2025-01-13
toc: true
---

::: {.callout-note}
This template follows [lecture 2.2 slides](../lecture-slides/lecture2.2-distributions-slides.qmd){target="_blank}.
:::

## Setup

```{r}
#| eval: true
#| echo: true
#| message: false
#| warning: false
#..........................load packages.........................
library(tidyverse)
library(chron)
library(naniar)

#..........................import data...........................
mko <- read_csv("https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-sbc.2007.17&entityid=02629ecc08a536972dec021f662428aa")

#..........................wrangle data..........................
mko_clean <- mko |>

  # keep only necessary columns ----
  select(year, month, day, decimal_time, Temp_bot, Temp_top, Temp_mid) |>

  # create datetime column (not totally necessary for our plots, but it can be helpful to know how to do this!) ----
  unite(date, year, month, day, sep = "-", remove = FALSE) |>
  mutate(time = chron::times(decimal_time)) |>
  unite(date_time, date, time, sep = " ") |>

  # coerce data types (see https://www.neonscience.org/resources/learning-hub/tutorials/dc-convert-date-time-posix-r for overview of POSIXct vs POSIXlt) ----
  mutate(date_time = as.POSIXct(date_time, "%Y-%m-%d %H:%M:%S", tz = "GMT"), 
         year = as.factor(year),
         month = as.factor(month),
         day = as.numeric(day)) |>

  # add month name by indexing the built-in `month.name` vector ----
  mutate(month_name = as.factor(month.name[month])) |>

  # replace 9999s with NAs ----
  naniar::replace_with_na(replace = list(Temp_bot = 9999, 
                                         Temp_top = 9999, 
                                         Temp_mid = 9999)) |>

  # select/reorder desired columns ----
  select(date_time, year, month, day, month_name, Temp_bot, Temp_mid, Temp_top)

#......................explore missing data......................

# counts and percentage of missing data by year ----
see_NAs <- mko_clean |> 
  group_by(year) |> 
  naniar::miss_var_summary() |>
  filter(variable == "Temp_bot")

# visualize missing Temp_bot ----
bottom <- mko_clean |> select(Temp_bot)
missing_temps <- naniar::vis_miss(bottom)
```

## Histograms

- represent distribution of a numeric variable(s), which is cut into several bins -- height of bar represents # of observations in that bin

### Too many groups

*Note the message, to remind us to consider adjusting our `binwidth`* 

```{r}
#| eval: true
#| echo: true

# histogram with all 12 months ----
```

### Alt 1: small multiples

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# histogram faceted by month ----
```

### Alt 2: fewer groups + update colors + modify bin widths

```{r}
#| eval: true
#| echo: true
#| warning: false

# histogram with fewer months ----
mko_clean |> 
  mutate(month_name = factor(month_name, levels = month.name))
```

## Density plots

- represent data distribution of a numeric variable(s); uses KDE to show probability density function of the variable; where the area under the curve is equal to 1

### Too many groups

```{r}
#| eval: true
#| echo: true
#| warning: false

# density plot with all 12 months ----
mko_clean |> 
  mutate(month_name = factor(x = month_name, levels = month.name)) 
```

### Alt 1: small multiples

```{r}
#| eval: true
#| echo: true
#| warning: false

# density plot faceted by month ---- 
mko_clean |> 
  mutate(month_name = factor(month_name, levels = month.name)) |> 
  ggplot(aes(x = Temp_bot)) +
  geom_density() 
```

### Alt 2: fewer groups + update colors + modify band widths

```{r}
#| eval: true
#| echo: true
#| warning: false

# density plot with fewer months ----
mko_clean |> 
  filter(month_name %in% c("April", "June", "October")) |> 
  ggplot(aes(x = Temp_bot)) +
  geom_density() 
```

## A few more histograms & density plots

### Distinction: histograms vs. density plots

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

# create some dummy data ----
dummy_data <- data.frame(value = c(rnorm(n = 100, mean = 5),
                                   rnorm(n = 200, mean = 10)),
                         group = rep(c("A", "B"),
                                     times = c(100, 200)))

# histogram ----
ggplot(dummy_data, aes(x = value, fill = group)) +
  geom_histogram(position = "identity", alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.75)

# density plot ----
ggplot(dummy_data, aes(x = value, fill = group)) +
  geom_density(alpha = 0.7) +
  geom_rug(aes(color = group), alpha = 0.75)
```

### Combining geoms

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# histogram + density plot ----
```

### Compare groups to a whole

```{r}
#| eval: true
#| echo: true
#| warning: false
library(palmerpenguins)

# use `after_stat(count)` to plot density of observations ----
ggplot(penguins, aes(x = body_mass_g, y = after_stat(count))) +
 
  # plot full distribution curve with label "all penguins"; remove 'species' col so that this doesn't get faceted later on ----
  geom_density(data = select(penguins, -species), 
               aes(fill = "all penguins"), color = "transparent") +
  
  # plot second curve with label "species" ----
  geom_density(aes(fill = "species"), color = "transparent") +
  
  # facet wrap by species ----
  facet_wrap(~species, nrow = 1) +
  
  # update colors, x-axis label, legend position ----
  scale_fill_manual(values = c("grey","#0C8346"), name = NULL) +
  labs(x = "Body Mass (g)") +
  theme(legend.position = "top")
```

## Ridgeline plots

- show distribution of a numeric variable for multiple groups

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#| layout-ncol: 2

# basic ridgeline plot ----

# fill with color gradient ----

```

### Alt 1: reorder groups + adjust overlap & tails

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# ridgeline plot with reordered months ----
ggplot(mko_clean, aes(x = Temp_bot, y = month_name, fill = after_stat(x))) +
  ggridges::geom_density_ridges_gradient() +
  scale_fill_gradientn(colors = c("#2C5374","#849BB4", "#D9E7EC", "#EF8080", "#8B3A3A"))
```

### Alt 2: add quantiles

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false

# ridgeline plot with quantiles ----
ggplot(mko_clean, aes(x = Temp_bot, y = month_name))
```

### Alt 3: jitter raw data

```{r}
#| eval: true
#| echo: true
#| layout-col: 2
#| warning: false
#| message: false

# jittered points ----

# raincloud ----

```

## Boxplots

- *summarize* the distribution of a numeric variable for one or several groups

```{r}
#| eval: true
#| echo: true
#| warning: false

# boxplot with all 12 months ----
ggplot(mko_clean, aes(x = month_name, y = Temp_bot)) +
  geom_boxplot() +
  scale_x_discrete(limits = rev(month.name)) +
  coord_flip()
```

### Alt 1: modify outliers

```{r}
#| eval: true
#| echo: true
#| warning: false

# boxplot with modified outliers ----
```

### Alt 2: hightlight a group

```{r}
#| eval: true
#| echo: true
#| warning: false

# highlight a particular group ----
```

### Alt 3: jitter raw data (using `{palmerpenguins}` data) 

```{r}
#| eval: true
#| echo: true
#| warning: false

# add jittered data points ----
```

### Alt 4: dodged groups

```{r}
#| eval: true
#| echo: true
#| warning: false

# dodge groups by year ----
```

### Alt 5: overlay beeswarm

```{r}
#| eval: true
#| echo: true
#| warning: false

# boxplot + beeswarm ----
ggplot(na.omit(penguins), aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  coord_flip() 
```

## Violin plots

- visualize distribution of a numeric variable for one or several groups; great for multiple groups with lots of data

```{r}
#| eval: true
#| echo: true
#| warning: false

# violin plot ----
ggplot(mko_clean, aes(x = month_name, y = Temp_bot))
```

### Alt 1: overlay boxplot

```{r}
#| eval: true
#| echo: true
#| warning: false

# violin + boxplot ----
# ~ copy code from previous chunk ~
```

### Alt 2: half-violin half-dot plot

```{r}
#| eval: true
#| echo: true
#| warning: false

# half violin + half dot plot ----
```

